The need for accuracy and smoothness in numerical simulations

### 2024-09-08 slides/pristina/cckm-fshmn-2024.tex
Adding the slides for Pristina presentation

\date{PPAM 2024, September 8-11th 2024,  Ostrava, Czechia}

@@

  This is not a talk about, say, 
  \vspace{0.5cm}
\begin{enumerate}\setlength\itemsep{1em}
\item computational speed
\item parallelism
\item energy efficency
\end{enumerate}
\vspace{0.5cm}
This is a talk about
\vspace{0.5cm}
\begin{center}
  \underline{the relevance of the numbers returned by our simulations}
\end{center}
}

\frame{\frametitle{A simple experiment}

  We did two sets of simulations of a molecule in water
  \vspace{1cm}
  \begin{enumerate} \setlength\itemsep{1em}
    \item fixed real time
    \item two different force-fields: \texttt{charmm36}, \texttt{oopsaal}
    \item identical initial conditions
    \item ran the constraint solver to stagnation
   \end{enumerate}
  \vspace{1cm}
  and tracked the final energy as a function of the number of steps.

\frame{% \frametitle{Warning: Disturbing pictures below}

 \begin{figure}
    \includegraphics[width=10.5cm]{oplsaaltol12.pdf}
  \end{figure}

  \begin{figure}
    \includegraphics[width=10.5cm]{charmm36tol12.pdf}
  \end{figure}

%% \frame{\frametitle{External ballistics}

%%   \includegraphics[width=12cm]{range_rk1_mwe2.pdf}

%%   The total force $\bm{F}$ acting on the shell 
%%   \begin{equation}
%%     \bm{F} = m \bm{g} - \frac{1}{2} \rho(y) A C_D(\nu,y) \|\bm{v} - \bm{w}\|_2 ( \bm{v} - \bm{w} )
%%   \end{equation}
%%   is the combination of gravity and aerodynamic drag.
%% }

%% \frame{\frametitle{External ballistics}

%% \begin{block}{Key questions}
%%   Does it matter
%%   \begin{enumerate}
%%   \item if the drag coefficient
%%     \begin{equation*}
%%       \nu \rightarrow C_D(\nu)
%%     \end{equation*}
%%     is smooth or not?
%%   \item if the event equation
%%     \begin{equation*}
%%       y(t) = 0
%%     \end{equation*}
%%     is solved accurately or not?
%%       \end{enumerate}
%%  \end{block}
%% }

@@

  If one of the following is true:
  \vspace{0.5cm}
  \begin{enumerate}
  \item the central functions are not smooth enough
  \item the central equations are not solved accurately enough
  \end{enumerate}
  \vspace{0.5cm}
  then we may lose the ability to
  \vspace{0.5cm}
  \begin{enumerate}
  \item assert that rounding errors are irrelevant
  \item estimate the discretization error
  \item compare the model against the physical reality
  \end{enumerate}
  \vspace{1cm}
}

%% \frame{\frametitle{The secondary point of this talk}
%%   It might be possible to assert experimentally if
%%   \vspace{0.5cm}
%%   \begin{enumerate} \setlength\itemsep{1em}
%%   \item your functions are sufficiently smooth
%%   \item your equations are solved with sufficient accuracy
%%   \end{enumerate}
%% }

@@

  We validate our models by demonstrating that
  \begin{equation}
    P - T \approx 0
  \end{equation}
  is a good approximation, but
  \begin{equation}
    P - T \approx P - \hat{A}_h
  \end{equation}
  is not necessarily a good approximation. We have
  \begin{equation}
    P - T = (P - \hat{A}_h) - (T - A_h) - (A_h - \hat{A}_h)
  \end{equation}
  so we need to assert that
  \begin{align}
    |T- A_h| & \ll |P - \hat{A}_h| \\
    |A_h - \hat{A}_h| & \ll |P - \hat{A}_h|
  \end{align}
}

@@

  The key is to have an asymptotic error expansion (AEX)
  \begin{equation}
    E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
  \end{equation}
  where
  \begin{equation}
    0 < p < q < r
  \end{equation}
  are not necessarily integers.
}

\frame{\frametitle{Elementary results}

Define 
\begin{equation}
  R_h := \frac{A_h - A_{2h}}{2^p - 1}, \quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
\end{equation}
Then
\begin{equation}
  \frac{E_h - R_h}{h^q} \rightarrow \text{constant}
\end{equation}
and
\begin{equation}
  \frac{2^p - F_h}{h^{q-p}} \rightarrow \text{constant}
\end{equation}
which implies
\begin{equation}
  \log | 2^p - F_h| \approx \log|\text{constant}| + (q-p) \log h
\end{equation}
}

\frame{\frametitle{An example from elementary numerical analysis}

@@

@@

  The total force on an ion is
  \begin{equation}
    \bm{F}(\bm{r_i}) =  - \alpha \sum_{j \not = i} \frac{1}{\|\bm{r}_i - \bm{r}_j\|^3}(\bm{r}_i - \bm{r}_j) - \beta \bm{r_i} - \gamma \bm{v}_i
  \end{equation}
  We measure the total kinetic energy $A_h$ at a fixed time as function of the stepsize $h_k = 2^{-k} h_0$. 
}

@@

\frame{\frametitle{Why should this concern the PPAM community?}

@@

>>>>

\date{Visit to FSHMN \\ Pristina, Kosovo, September 12th 2024}

@@

  \begin{itemize}
  \item Danish nationality
  \item Ass. Prof. of Comp. Sci. at Ume{\aa} University, Sweden
  \item Ph.D in mathematics from Purdue University, Indiana, USA
  \item MSc. in mathematics from Aarhus University, Denmark
  \item Research interests
    \begin{itemize}
    \item Structured linear system
    \item Nonsymmetric eigenvalue problems
    \item Solution of nonlinear constraint equations
    \item High performance scientific computing
    \end{itemize}
  \item Software
    \begin{itemize}
    \item Coauthor of StarNEIG
    \item Coauthor of ILVES
    \end{itemize}
  \end{itemize}
}

%% \frame{\frametitle{A simple experiment}

%%   We did two sets of simulations of a molecule in water
%%   \vspace{1cm}
%%   \begin{enumerate} \setlength\itemsep{1em}
%%     \item fixed real time
%%     \item two different force-fields: \texttt{charmm36}, \texttt{oopsaal}
%%     \item identical initial conditions
%%     \item ran the constraint solver to stagnation
%%    \end{enumerate}
%%   \vspace{1cm}
%%   and tracked the final energy as a function of the number of steps.

%% }

%% \frame{% \frametitle{Warning: Disturbing pictures below}

%%  \begin{figure}
%%     \includegraphics[width=10.5cm]{oplsaaltol12.pdf}
%%   \end{figure}

%%   \begin{figure}
%%     \includegraphics[width=10.5cm]{charmm36tol12.pdf}
%%   \end{figure}

%% }

\begin{block}{Key questions}
  Does it matter
  \begin{enumerate}
  \item if the drag coefficient
    \begin{equation*}
      \nu \rightarrow C_D(\nu)
    \end{equation*}
    is smooth or not?
  \item if the event equation
    \begin{equation*}
      y(t) = 0
    \end{equation*}
    is solved accurately or not?
      \end{enumerate}
 \end{block}
}

@@

  If one of the following is true:
  \vspace{0.5cm}
  \begin{enumerate}
  \item the central functions are not smooth enough
  \item the central equations are not solved accurately enough
  \end{enumerate}
  \vspace{0.5cm}
  then we will almost certainly lose the ability to
  \vspace{0.5cm}
  \begin{enumerate}
  \item assert that \underline{rounding} errors are irrelevant
  \item estimate the \underline{discretization} error
  \item estimate the \underline{modelling} error
  \end{enumerate}
  \vspace{1cm}
}

@@

  We validate our models by demonstrating that
  \begin{equation}
    P - T \approx 0
  \end{equation}
  is a good approximation, but
  \begin{equation}
    P - T \approx P - \hat{A}_h
  \end{equation}
  is not necessarily a good approximation. We have
  \begin{equation}
    \underset{\text{modelling error}}{\underbrace{P - T}} = (P - \hat{A}_h) - \underset{\text{discretization error}}{\underbrace{(T - A_h)}} - \underset{\text{computational error}}{\underbrace{(A_h - \hat{A}_h)}}
  \end{equation}
  so we need to assert that
  \begin{align}
     |T- A_h| & \ll |P - \hat{A}_h| \\
     |A_h - \hat{A}_h| & \ll |P - \hat{A}_h|
  \end{align}
}

@@

  The key is to have an asymptotic error expansion (AEX)
  \begin{equation}
    E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
  \end{equation}
  where
  \begin{equation}
    0 < p < q < r
  \end{equation}
  are not necessarily integers and $\alpha$, $\beta$ are independent of $h$.
}

\frame{\frametitle{Basic definitions}

We define Richardson's error estimate by
\begin{equation}
  R_h := \frac{A_h - A_{2h}}{2^p - 1}
\end{equation}
and Richardson's fraction by
\begin{equation}
\quad F_h:= \frac{A_{2h} - A_{4h}}{A_h - A_{2h}} 
\end{equation}
}

\frame{\frametitle{Elementary results}
  If
 \begin{equation}
    E_h = T - A_h = \alpha h^p + \beta h^q + O(h^r), \quad h \rightarrow 0_+
  \end{equation}
 then
 \begin{equation}
  \frac{E_h - R_h}{h^q} \rightarrow \text{constant}
\end{equation}
 and
\begin{equation}
  \frac{2^p - F_h}{h^{q-p}} \rightarrow \text{constant}
\end{equation}
}

\frame{\frametitle{Implications}

  \begin{enumerate}
  \item We can determine $p$ from
    \begin{equation}
      F_h \rightarrow 2^p
    \end{equation}
  \item We can determine $q$ from
    \begin{equation}
      \log | 2^p - F_h| \approx \log|\text{constant}| + (q-p) \log h
    \end{equation}
  \item We can then compute $R_h$ and estimate
    \begin{equation}
      E_h \approx R_h
    \end{equation}
    when $h$ is \textit{sufficiently} small.
  \end{enumerate}
}

\frame{\frametitle{Numerical integration \texttt{rint\_mwe1}}

@@

\frame{\frametitle{Numerical integration \texttt{rint\_mwe2}}

  Our target value is 
  \begin{equation}
    T = \int_0^1 f(x) dx, \quad f(x) = \sqrt{x}.
  \end{equation}
  Our approximation is the composite trapezoidal rule
  \begin{equation}
    A_h = \frac{h}{2} \sum_{j=0}^{n-1} \left( f(x_i) + f(x_{i+1}) \right), \quad x_i = ih, \quad nh = 1
  \end{equation}
  We compute $A_h$ for
  \begin{equation}
    h = h_k = 2^{-k}
  \end{equation}
}

  \begin{figure}[ht]
        \begin{minipage}[b]{0.45\linewidth}
            \centering
            \includegraphics[width=\textwidth]{rint_mwe2a.pdf}
            \caption{The evolution of $F_h$}
            \label{fig:a}
        \end{minipage}
        \hspace{0.5cm}
        \begin{minipage}[b]{0.45\linewidth}
            \centering
            \includegraphics[width=\textwidth]{rint_mwe2b.pdf}
            \caption{The accuracy of $R_h$}
            \label{fig:b}
        \end{minipage}
    \end{figure}
  We observe that
  \begin{enumerate}
  \item the asymptotic range is much wider (good!)
  \item the error estimate is less accurate (mostly harmless)
  \end{enumerate}

@@

  The total force on an ion is given
  \begin{equation}
    \bm{F}(\bm{r_i}) =  - \alpha \sum_{j \not = i} \frac{1}{\|\bm{r}_i - \bm{r}_j\|^3}(\bm{r}_i - \bm{r}_j) - \beta \bm{r_i} - \gamma \bm{v}_i
  \end{equation}
  where $r_k$ is the position of the $k$th ion and $v_k$ is it velocity.
  \vspace{1cm}
  \begin{itemize}
  \item We wish to know the total kinetic energy $T$ at a fixed time.
  \item We compute approximation $A_{h_k}$ for $h_k = 2^{-k} h_0$.
  \end{itemize}
}

@@

\frame{\frametitle{Why should this concern the computational scientist?}

@@

\frame{\frametitle{Links}

  \begin{figure}
    \includegraphics[width=6cm]{homepage.png}
  \end{figure}
  \begin{center}
    Thank your for your attention
  \end{center}
}
